Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-04-07 demystify.tex
updates
typos

\begin{commentbox}[Vu:]
  One of the reasons I create this document is that my colleagues at GMU are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS PhD program receives about 400 PhD applications, most of which are international but only 4--5 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome.
\end{commentbox}

@@

While the information presented here can be useful for all students interested in applying to a CS PhD program in the US, this guideline is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (similarly, for domestic students from universities with no PhD programs or limited research experiences). 
Students from top schools with strong research cultural and experiences might already know most of the information presented here.
My goal is to level the playing field by providing information that is not readily available to less privileged students.  I also hope to encourage more students from these countries to apply to CS PhD programs in the US.

@@

After you submit your PhD application (usually around December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screen process is done through a central university system, i.e., not by faculty from the CS department. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depends on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that most large schools, including GMU, have separate adcoms for MS programs.

@@

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. Some members are tenured faculty. The chair of the committee will likely a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

At GMU, we usually decide that a full-time PhD candidate is either (i) admit with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding, or we reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak
LORs.

@@

\paragraph{Why CS depts do not waive application fee?}  Some universities do waive, e.g., Rice for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not.  In any case, application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. In fact, a strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).
However, obtaining letters from famous people
can be challenging for international students, who might not have much interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) that know you well enough to talk about \emph{your research experience and capabilities}. Many students get letters from supervisors from companies where they did internships or are working at. It is OK as long as it is a research-based personalized letter (once again, we are talking about PhD applications, not MS). Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potentials. Letters focusing your course work or non-research projects at a company won't carry much weight.

\begin{commentbox}[Vu:]
  If your grading system is not US standard or you are from a good school unknown outside your country, you can ask your reference writers to explain about that in their letters.  For example, ``Bach Khoa'' are the top universities in Vietnam for STEM studies but few people outside Vietnam know about them.  So if you are from there, you should ask your reference writers to mention that.
\end{commentbox}

\paragraph{How to help your LOR writer} Your LoR writer can ask you to provide additional information to help them write your letter.  For example, they might ask your GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and can complement that with their own perspective.  It might also be a good idea to provide them such information even when they do not ask, especially if you have not interacted with them much or have not done much research with them.

\paragraph{Self-written Letters} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have little values and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?.  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter).

\paragraph{Generic Letters} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well ask me to write letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

There are various guides on writing SOP , and \href{https://cs-sop.org/}{many example statements} are available. In short, focus on your research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidences. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..). Also, do not use more than 2 pages (you can have references in the third page).

@@

\begin{commentbox}[Vu:]
  I often read LORs and SOP first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something really standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or  adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to wrong schools or mixing
  facts (e.g., talking about school X but mentioned about working with
  profs. at school Y; and definitely do not talk about George Washington when applying to George Mason). I have seen such statements more time that I
  should.
\end{commentbox}

@@

\paragraph{Publications} A concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has successfully involved in research.  Applicants  admitted to top schools often have multiple first-authored papers at top places.

\begin{commentbox}[Vu:]
  Many international students mention Scopus Q1, which consists of various journals from IEEE, Elsevier, and many other publishers.  I don't know/recognize many of journals listed in Scopus Q1. This might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed there, so devote some part in your statement to discuss the significance of your papers.
\end{commentbox}

However, it is understandable that many students do not have the opportunities to publish papers. Thus, general writings, even those under submissions or rejected, would still be good (much better than nothing).  Be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still upload them, mention them in your statement, and explain why they are good.

@@

\paragraph{Work Experiences} Work experiences at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicit say something about them in your statement.

@@

\begin{commentbox}[Thanh:]
  Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?.
  \tcblower
  Vu: One way is looking at what well-known researchers publish at. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  Sometime PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members in their CS PhD adcom ask me to evaluate applicants from Vietnam.  During my time at UNL and now here at GMU, I have looked at Vietnamese applications (whether they are assigned to me or not) and provide input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}

@@

\begin{commentbox}[Thanh:]
  Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, an 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
  \tcblower
  \textbf{Vu}: Vietnamese students and even faculty often lament how this grading system hurts Vietnamese students applying abroad. One way to mitigate this is making these issues known in your SOP.  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Universities with Vietnamese profs} are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

\paragraph{English Tests} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test.

\textbf{Advice}: Just do well enough to pass minimum requirement set by the university. %, which is required for TA (\autoref{sec:funding}). 
Just as with grades (\autoref{sec:your-school}) and GRE, having a high scores in these tests might not help, but having too low scores can be a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\textbf{Advice}: do not spend much time talking about non-research related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab, e.g., Microsoft, DoE, VinAI in Vietnam, etc (\autoref{sec:research-experiences}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potentials in their LOR (\autoref{sec:lor}).

@@

\textbf{Advice}: Prepare your CV in such away that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, Teaching Experience.

@@

After you submitted your applications, the waiting game begins! For many students, this is a very stressful time. This section provides some information and tips to help you get through this time.

@@

\paragraph{Funding \autoref{sec:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and permanent residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships can be very competitive and will improve your admission chance if you get them.

>>>>

\begin{commentbox}[Vu:]
  One of the reasons I created this document is that my colleagues at GMU are interested in recruiting Vietnamese students and are surprised when seeing very few applications in Vietnam (see \autoref{sec:ack}). Each year our CS PhD program receives about 400 PhD applications, most of which are international but only 4--5 are from Vietnam. In general the number of CS PhD applications from Vietnam to US universities is very low and more would be welcome.
\end{commentbox}

@@

While the information presented here can be useful for all students interested in applying to a CS PhD program in the US, this guideline is primarily intended for \textbf{international students} from \textbf{smaller countries and less well-known universities} (similarly, for domestic students from universities with no PhD programs or limited research experiences). 
Students from top schools with strong research cultures and experiences might already know most of the information presented here.
My goal is to level the playing field by providing information that is not readily available to less privileged students.  I also hope to encourage more students from these countries to apply to CS PhD programs in the US.

@@

After you submit your PhD application (usually around December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screening process is done through a central university system, i.e., not by faculty from the CS department. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depend on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that most large schools, including GMU, have separate adcoms for MS programs.

@@

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. Some members are tenured faculty. The chair of the committee will likely be a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

At GMU, we usually decide that a full-time PhD candidate is either (i) admitted with funding (\autoref{sec:funding}) or (ii) rejected. In other words, in most cases, we either
admit you with full funding or reject your application. In some rare cases, we admit
without funding because you have funding on your own, e.g.,
supported by your government or having external fellowships. We justify
our decision with a summary of your application, where we list
strengths, e.g., a well-known school, and weaknesses, e.g., weak
LORs.

@@

\paragraph{Why CS depts do not waive the application fee?}  Some universities do waive, e.g., Rice for all PhD applicants and many universities for domestic students (\autoref{sec:domestic-students}), however most do not.  In any case, the application fee is typically a requirement of the university. Individual departments and programs do not have the flexibility to waive the application fee, even if they want to.

@@

Having a strong letter from an internationally recognized researcher will \emph{greatly strengthen} your application. A strong letter from a well-known researcher can outweigh the lack of publications (or low GPA or unknown school).
However, obtaining letters from famous people
can be challenging for international students, who might not have many interactions with such experts. So it is fine to have a letter from people (e.g., profs, researchers, postdocs who mentored you) who know you well enough to talk about \emph{your research experience and capabilities}.

Many students get letters from supervisors from companies where they did internships or are working.  It is OK as long as it is a research-based personalized letter. Again, the emphasis here is \emph{research}, i.e., the letters should describe your research experiences and potential. Letters focusing on your course work or non-research projects at a company won't carry much weight.

\begin{commentbox}[Vu:]
  If your grading system is not US standard or you are from a good school unknown outside your country, you can ask your reference writers to explain that in their letters.  For example, ``Bach Khoa'' are the top universities in Vietnam for STEM studies but few people outside Vietnam know about them.  So if you are from there, you should ask your reference writers to mention that.
\end{commentbox}

\paragraph{How to help your LOR writer} Your LoR writer can ask you to provide additional information to help them write your letter.  For example, they might ask about GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and can complement that with their own perspective.  It might also be a good idea to provide them with such information even when they do not ask, especially if you have not interacted with them much or have not done much research with them.

\paragraph{Self-written Letters} Many international students write letters themselves, typically due to the requests of their profs., and have them signed by their profs. Such letters have little value and are considered weak by reviewers---why can you not even find someone who cares or knows enough about you to write a candid personal reference letter?  Instead of the reference writer talking about you, in this case it is you who write about yourself (and they just sign the letter).

\paragraph{Generic Letters} When the writers do not know much about the applicants (e.g., just taking some course with them or not making any impression to write about), they might write a generic and short letter, which is not useful and also considered weak.  It might be a good idea to directly ask if the prof. is willing to write a \emph{strong} letter for you. If not, then you should ask someone else.  For example, if a student I don't know well asks me to write a letter for them, I will explicitly tell them I don't know them that well to write much about them, and such a short, generic, and weak letter will not help their case.

@@

There are various guides on writing SOP , and \href{https://cs-sop.org/}{many example statements} are available. In short, focus on your research motivation and convince us that you can achieve it through your experience, e.g., published papers. You should also back up your claims with concrete evidence. For example, if you say you have experience with teaching, then show what you did, e.g., undergrad TA or mentoring someone.  If you say you work on project X, then show some results (paper submitted (or even rejected), achieved X\% performance improvement over ..). Also, do not use more than 2 pages (you can have references on the third page).

@@

\begin{commentbox}[Vu:]
  I often read LORs and SOP first. If I am
  persuaded by then, I would skim over other factors and advocate for
  admission (unless I see red flags in other parts). However, if I am not
  convinced, then I will likely recommend rejection (unless I see
  something really standout in other parts).
  \tcblower
  Do careful research on professors, don't mention \emph{emeritus} or adjunct faculty (see more about various types of faculty in \autoref{sec:faculty-types}).
  Also, be careful not to send statements to the wrong schools or mix
  facts (e.g., talking about school X but mentioning working with
  profs. at school Y; and do not talk about George Washington when applying to George Mason). I have seen such statements more times than I should.
\end{commentbox}

@@

\paragraph{Publications} A concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the applicant has been successfully involved in research.  Applicants admitted to top schools often have multiple first-authored papers at top places.

\begin{commentbox}[Vu:]
  Many international students mention Scopus Q1, which consists of various journals from IEEE, Elsevier, and many other publishers.  I don't know/recognize many of the journals listed in Scopus Q1. This might be something to be mindful of, as \textbf{CS} faculty might not be too familiar with Scopus or journals listed there, so devote some part in your statement to discuss the significance of your papers.
\end{commentbox}

However, it is understandable that many students do not have the opportunity to publish papers. Thus, general writings, even those under submissions or rejected, would still be good (much better than nothing).  Be sure to upload your papers with your application and talk about them in your statement (see \autoref{sec:research-statement}).  Note that local conferences and non-English journals or conferences do
not carry as much weight since their quality is often unknown to US faculty. However, if you have published in such places, you should still upload them, mention them in your statement, and explain why they are good.

@@

\paragraph{Work Experiences} Work experiences at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicitly say something about them in your statement.

@@

\begin{commentbox}[Thanh:]
  Due to academic culture, professors in Vietnam usually aim for (international) journals instead of conferences. Could you give some tips on how to know whether a journal is good (CSrankings, unfortunately, only consider conferences)?
  \tcblower
  Vu: One way is looking at what well-known researchers publish. For example, if you are interested in a field X, you can use CSRankings to look at active faculty in X, and then look at their websites to see what journals they publish at.
\end{commentbox}

@@

\begin{commentbox}[Vu:]
  Sometimes PhD adcom in the US will share a document such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Foreign-Top-Schools}{this one}, which lists the top schools in several countries. We also ask other faculty and students if we think they know about the place.  For example, when I was a postdoc at UMD, members of their CS PhD adcom asked me to evaluate applicants from Vietnam.  During my time at UNL and now here at GMU, I have looked at Vietnamese applications (whether they are assigned to me or not) and provided input to their reviewers, e.g., X is the top tech school in Vietnam and so it should be \emph{top} instead of \emph{unknown} foreign, which makes a huge difference.
\end{commentbox}
\begin{commentbox}[Deepak:]
  If an applicant is anxious about their school not being known outside their country, they can provide information about their school and department, with independent sources where such information can be verified.
\end{commentbox}

@@

\begin{commentbox}[Thanh:]
  Vietnamese universities typically offer specialized programs, such as the talented engineer program at HUST, that have highly competitive entrance exams and a limited number of available slots (e.g., 30 per year). However, these programs often set higher requirements for students, including more demanding tests and assignments, resulting in lower GPAs and overall rankings. For example, a 3.5 GPA students from such talented programs are typically much better than a 4.0 GPA students not in those programs.  Similarly, variations in GPA standards exist among different universities, with technical universities generally having lower GPAs than economical universities. These make gaining admission in the US difficult as US faculty are not familiar with these issues.
  \tcblower
  \textbf{Vu}: Vietnamese students and even faculty often lament how this grading system hurts Vietnamese students applying abroad. One way to mitigate this is making these issues known in your SOP.  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Universities with Vietnamese profs} are probably aware of them, but in general your letter writers \emph{and} you can explicitly mention these in their letters and your statement.
\end{commentbox}

@@

\paragraph{English Tests} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English tests.

\textbf{Advice}: Just do well enough to pass the minimum requirement set by the university. %, which is required for TA (\autoref{sec:funding}). 
Just as with grades (\autoref{sec:your-school}) and GRE, having high scores in these tests might not help, but having too low scores can be a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\textbf{Advice}: do not spend much time talking about non-research-related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab, e.g., Microsoft, DoE, VinAI in Vietnam, etc (\autoref{sec:research-experiences}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potentials in their LOR (\autoref{sec:lor}).

@@

\textbf{Advice}: Prepare your CV in such away that it allows the reviewers to quickly scan to identify major achievements, e.g., Publications, Programming Competition Awards, and Teaching Experience.

@@

After you submit your applications, the waiting game begins! For many students, this is a very stressful time. This section provides some information and tips to help you get through this time.

@@

\paragraph{Funding \autoref{sec:funding}} You have more opportunities for funding, e.g., through government scholarships for US citizens and permanent residents.  You can also apply to specific programs \emph{before} you start your PhD, e.g., NSF Graduate Research Fellowship Program (GRFP) and Hertz Foundation Fellowship.  These fellowships are very competitive and can significantly improve your admission chance if you get them.
